## Intro-to-R 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Since some of you may not have any experience using R, we will very quickly go through some key concepts to help you understand better what we will doing in the analysis section later. 

### Mathematical operations 

Type and test these operations in the console in Rstudio 

Addition

```{r}

```

Subtraction 
```{r}

```


Multiplication 
```{r}

```

Division
```{r}

```

Modulo 
```{r}

```

Square
```{r}

```




### Comparison

Greater than
```{r}

```

Less than 
```{r}

```

Equals 
```{r}

```

Does not equal 
```{r}

```

### Basic types 

There are 6 basic data types in R: logical, numeric, integer, character, raw and complex. Today (and in general) only the first four are important. 

```logical``` Can only have two values ```TRUE``` or ```FALSE``` (Shorthand:```T```,```F```). 

```numeric``` All numbers with or without decimal ```100.34```

```integer``` Whole numbers ```123L``` A number appended with L suffix denotes an integer in R

```character``` Character or string values. Enclosed in either single 'microbe' or double "stroke" quotes. 

```{r}
class(TRUE)
class(3.0)
class(6L)
class("Hello")
```

### Variable assignment 

Variables are assigned in R using the ```<-``` operator. A variable can be though of a a "box" you store a value in for later use.

They can be used to store a value which you want to be constant throughout your code e.g. a file name 
```{r}

```
Or to save time by storing long strings/numbers 

```{r}

```

### Data structures 

Four most commonly used data structures 

#### 1D
Vectors - ordered collection of data of a given length and all of the same type:

```{r}

```

Lists - ordered collection of data/objects of a given length. Can be composed of heterogenous types. 

```{r}

```

#### 2D 

Matrices - 2D data structure (rows x columns) of the same type

```{r}

```

Dataframes - 2D heterogenous data structure (rows x columns)
df) 

```{r}

```

Objects such as vectors and dataframes can also be stored in variables as you saw above. ```df``` and ```vector2``` are both variables. 

### Control flow

As well as writing individual statements like we have done above we can also use logic to control the flow of our code, allowing us to repeat bits of code or only run if a given condition is met. 

for loops allow us to repeat code a specified number of times e.g.

```{r}

```

This prints out each element in the ```vector2```variable we defined earlier. Not particularly interesting though.. 
We will discover a more interesting use case later. 

if/else statements allow us to control the flow of our code better: 

```{r}

```

Change the values of x and y and see how the output changes. 

### Functions 

Functions are used to abstract repetitive code in something which is reusable and modular, making code easier to read and debug. 

Using the above if else example we could create a function called ```greater_than``` which tells us which value is highest in a more modular way

```{r}

```

```{r}

```

More interesting functions might perform a calculation for us and return the value 

```{r}

```

Let's run this function on the matrix we created earlier to see what it does. 

```{r}

```

So far we have covered built-in functions and custom functions, but R has a huge 
open source library of packages called CRAN, as well as a specific repository
for bioinformatics, called bioconductor. 

### Loading external function 

When working with R, at some point you may find yourself requiring a specific 
function to perform some kind of analysis and plot data etc. It's usually the 
case that someone has already written a function for whatever you want to do

You can install packages to access functions written by others. 
External packages can be downloaded using the ```install.packages``` function 
and loaded using the ```library``` function. Somewhat annoyingly, bioconductor
packages are installed a little differently, and require you to first install
the "BiocManager" package and then install using ```biocManager::install```. They
can however be loaded as normal. 

If you need help with a function you can also type ?functionname in the console e.g. ?log10 and the help for that function will show up, detailing what the function does, what inputs it expects and what value(s) it returns. 

Once you get into writing your own functions, it's good practice to store them
in a separate R file and import them using the ```source``` function. 

### Data wrangling 

A few key concepts on loading and manipulating data. 

Reading data 

```{r}

```

Manipulating data 

Selecting columns 

```{r}

```

Filtering rows 

```{r}

```

Filtering and selecting in one with the pipe operator 

```{r}

```